---
title: "Question2"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.packages("ggplot2")
install.packages("tidyverse")
install.packages("janitor")
install.packages("ragg")
install.packages("ggsignif")
install.packages("broom")
library(ggplot2)
library(tidyverse)
library(janitor)
library(ragg)
library(ggsignif)
library(broom)
```

## QUESTION 2: Data Pipeline

# Introduction

I am interested in exploring the relationships between different factors in the palmer penguins dataset. I will first load, view and clean the raw data from the penguins dataset so that it is easier to use for data analysis. I have outlined by steps below.

First, I will load the palmer penguins dataset.

```{r loading palmerpenguins}
install.packages("palmerpenguins")
library(palmerpenguins)
```

I will save the raw data as a csv and create a folder named "data" in my working directory to store the data. Here I will keep the raw data, so I am not editing it. I will load the data from the saved version.

```{r save and load data}
dir.create("data")
write.csv(penguins_raw, "data/penguins_raw.csv") 
penguins_raw <- read.csv("data/penguins_raw.csv")
```

When I view the data, I can see that saving it as a csv put full stops in the titles where there were spaces.

```{r view a data, echo=TRUE}
head(penguins_raw)
names(penguins_raw)
```

I would like to clean this so I will define and call a function then save a clean version of the raw data. I am defining a function so that I can use it at a later date if needed. 

```{r defining a function}
clean_function <- function(penguins_data) {
  penguins_data %>%
    select(-Comments) %>% #removes comments column
    clean_names() %>% #cleans names
    na.omit()  %>% #remove NAs
    #shorten species names
    mutate(species = case_when(
      species == "Adelie Penguin (Pygoscelis adeliae)" ~ "Adelie",
      species == "Chinstrap penguin (Pygoscelis antarctica)" ~ "Chinstrap",
      species == "Gentoo penguin (Pygoscelis papua)" ~ "Gentoo"
    ))
}

```

I am calling the function and saving the now clean data in my "data" folder.

```{r  calling function}

penguins_clean <- clean_function(penguins_raw)

write.csv(penguins_clean, "data/penguins_clean.csv")

head(penguins_clean) #check if it is clean
```

# Hypothesis

Now that the data is clean, I can explore it. I am interested in whether there is a correlation between body mass and flipper length in each of the 3 species of penguins.

**My biological hypothesis is that there will be a positive correlation between body mass and flipper length in all 3 species because this ratio may be important to activities such as foraging, particularly in water.**

I will plot body mass against flipper length and group by Species. I have done this as a function so that it is easier to reuse and I can track edits.

```{r exploratory plot, echo=TRUE}
plot_exploratory_figure <- function(penguins_clean) {
  penguins_clean %>%
    ggplot(aes(x = body_mass_g, y = flipper_length_mm, colour = species)) +
    geom_point(alpha=0.5) +
    labs(title = "Body Mass Against Flipper Length for 3 Species of Penguin", x = "Body Mass (g)", y = "Flipper Length (mm)")
}

exploratory_figure <- plot_exploratory_figure(penguins_clean)

exploratory_figure
```

I will create a folder named "figures" and save my plot as a png here

```{r saving exploratory figure }
dir.create("figures")
agg_png("figures/exploratoryfigure.png", 
        width = 1000, height = 750, units = "px",res=300, scaling = 0.5)
exploratory_figure
dev.off()
```

# Stats method

From the exploratory figure, it appears that there may be a positive correlation between body mass and flipper length, in all 3 species. I now need to statistically test this. The **null hypothesis**, H0, will be that there is so significant relationship between body mass and flipper length. The **alternative hypothesis**, H1, will be that flipper length increases with body mass.

First, I will add a linear model to my plot

```{r add linear model}
exploratory_figure + geom_smooth(method="lm", formula = y~x)
```

It appears that each species has a positive correlation but they are not all the same. I will run my stats test by running a linear regression model for each species.

Adelie

```{r adelie linear regression model}
model_adelie <- lm(flipper_length_mm ~ body_mass_g, data = subset(penguins_clean, species == "Adelie"))
```

Chinstrap

```{r chinstrap linear regression model}
model_chinstrap <- lm(flipper_length_mm ~ body_mass_g, data = subset(penguins_clean, species == "Chinstrap"))
```

Gentoo

```{r gentoo linear regression model}
model_gentoo <- lm(flipper_length_mm ~ body_mass_g, data = subset(penguins_clean, species == "Gentoo"))

```

# Results

Here are the results of my linear regression models

```{r model results }
summary(model_adelie)
summary(model_chinstrap)
summary(model_gentoo)
```

I will add the p-values from the linear regression models results figure.

```{r results figure }

# I have used geom_smooth to add a linear model to my exploratory plot 
# I have annotated the plot with the results of the linear models: 
# Using annotate - this code tells r to go into the coefficient table of model and find the p value which is in row 2, column 4. Digits=3 specifies the digits. 

(results_figure <- 
  exploratory_figure +
   geom_smooth(method="lm", formula=y~x)+
  annotate("text", x = 6100, y = 180, hjust = 1, vjust = 0.5, label = sprintf("Adelie: p = %s", formatC(summary(model_adelie)$coef[2, 4],  digits = 3))) +
  annotate("text", x = 6100, y = 180, hjust = 1, vjust = 1.5,label = sprintf("Chinstrap: p = %s", formatC(summary(model_chinstrap)$coef[2, 4],  digits = 3))) +
  annotate("text", x = 6100, y = 180, hjust = 1, vjust = 2.5, label = sprintf("Gentoo: p = %s", formatC(summary(model_gentoo)$coef[2, 4], digits = 3))))

 
```

I will save this in my "figures" folder.

```{r save results figure}
agg_png("figures/resultsfigure.png", 
        width = 1000, height = 750, units = "px",res=300, scaling = 0.5)
results_figure
dev.off()
```

# Discussion

The results from the linear regression models allow us to reject the null hypothesis that there is no effect of body mass on flipper length. The intercept, coefficient and F-statistic p-values for Adelie, Chinstrap and Gentoo are all highly significant (p-value < 0.001). This suggests that for all 3 species, there is a positive relationship between body mass and flipper length and as body mass increases, flipper length increases. This means that body mass may be a predictor of flipper length but we cannot say that a larger body mass *causes* longer flippers.

The appropriateness of the model can also be questioned,if we look at the R-squared values, which were different. For Adelie, it was 0.2347, for Chinstrap 0.4107, and for Gentoo 0.5059. This means that the linear regression model explains more (around 51%) of the variance in flipper length for Gentoo penguins than for Adelie (around 23%) and Chinstrap (around 41%). This suggests that the relationship may not be perfectly linear and so, there may be other factors to consider which are also important.

# Conclusion

It can be concluded from a linear regression model that there may be a positive relationship between body mass and flipper length for all 3 species of penguin present in this data-set. As body mass increases, flipper length appears to increase. However, the relationship may not be linear and so it would be useful to run non-linear models. It would also be interesting to explore the biological factors that can explain this relationship.